From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-24 paper.tex
Global proof-reading (except sec 4.2)

``Engagement'', the cognitive, affective and behavioral state of interaction
with a computer application that ``makes the user want to be
there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
domains, and specifically in robotics, several variables and social signals have been proposed in the
literature to quantify it. A recent review of these is in~\cite{ivaldi2015towards}.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
with the current state of the interaction. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
connection events such as directed gaze, mutual facial gaze is proposed. Not
relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
a determining factor for the assessment. A recent study~\cite{anzalone} with
social robots in face-to-face scenarios proposed a set of metrics based on
non-verbal cues but manifesting some possible limitations in long-term
scenarios.

The variety of these approaches reflects the fact that \emph{engagement} remains
a broad concept, fairly ill-defined and thus difficult to operationalize.
Therefore, instead of introducing ``yet another metric of engagement'', we
propose in this paper to introduce the more specific concept of
``\emph{with-me-ness}'': to what extend the human remains ``with me'', the
robot, during our interaction.

We assess this level of ``with-me-ness'' by combining real-time estimation of the
attentional focus of the human with the expected, {\it a priori} focus
requested by the task, as decided by the robot.

As such, this article is organized in three sections: we first present a novel
method for on-line estimation of the focus of attention based on fast 6D head
pose estimation. We then present an extensive validation of this approach based on
a real-world field experiment with 6 children. We finally show how to compute in
real-time ``with-me-ness'' to provide a new \emph{in-the-moment} measurement of
the quality of interaction, and compare levels computed by the robot with a
ground truth based on manual post-hoc annotations.

@@

The relations between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction, and the related gaze patterns, has been part of classic textbooks
like~\cite{argyle1969social} since decades. As such, there is little doubt that
measuring the direction of gaze is a useful proxy to estimate the (visual) focus
of attention of a social agent, and indeed this is one of the basic tool used in
social psychology.

Estimating attention using gaze is certainly not new to robotics either. A
recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
way eye gaze research in HCI and social robotics, and is an essential read in
that regard. It remains however an active field of research, as illustrated by
several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
as attention tracking is essential in human-robot interaction

However, performing such a measure on a robot, in real-time, and in ecologically
valid environments (which rules out bulky or invasive apparatus like
eye-trackers, or techniques requiring fine calibration and/or static
interactions) remains a key challenge in the field of Human-Robot Interaction.

Looking at techniques that both operate on-line and have been deployed in field
experiments, one finds that most approaches rely on head pose estimation alone
(no eye gaze tracking) and are most often based on depth sensors (RGB-D) .
Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
and recent examples include~\cite{baxter2014tracking, anzalone}.

Approaches based on monocular 2D vision have been explored as
well~\cite{peters2010investigating}, with however limited robustness to
occlusions or lightning conditions, and over-reliance on tracking to maintain
real-time performances.  Our work relies on recent advances on template-based
face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) facial features extraction on 2D images, combined with 3D model
fitting, to obtain a fast, robust and stable 6D head pose estimate, that was
successfully deployed in field experiments involving child-robot interactions.

We then derive the field of attention from the head pose: according
to~\cite{stiefelhagen2002tracking} the head orientation's contribution in
overall gaze direction is 68.9\% showing agreement between both measurements.
In addition, it shows that the head orientation alone can get an average
accuracy of 88.7\% detecting the focus of attention in a meeting application
scenario using a head-mounted display with eye and head tracking.

While previous preliminary research in HRI seemed to indicate that deriving
attentional focus from head pose alone would not be accurate
enough~\cite{kennedy2015head}, we found in our case acceptable levels of
agreement between the robot observation and manual post-hoc annotations.

@@

As explained, we derive the visual field of attention from the head pose. Our
technique only involves a single monocular RGB camera for used for facial
feature extraction, and a static simplified 3D mesh of a human head
(Figure~\ref{head_pose}). 68 facial features are extracted using a fast
template-based face alignment algorithm by Kazemi and
Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
library\footnote{http://dlib.net}~\cite{dlib09}.  Eight of these features are
matched to their 3D counterparts and we rely on an iterative $PnP$ algorithm
(OpenCV's implementation) to compute the translation and rotation of the head
with respect to the camera frame. With this approach, knowing the intrinsic
parameters of the camera (calibrated camera) is required for an accurate
estimation of the absolute 3D localization of the head.

Besides being fast, the face alignment algorithm has been found to perform well
in terms of robustness, including in a range of difficult situations
(occlusions, head angle) as encountered in field experiments.
Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes with
one child.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{field_of_attention}
    \caption{\small The visual field of attention is approximated to a
        20$\degree$ cone, spanning from the head's sellion. The objects whose 3D
    pose intersect with this cone are considered \emph{in focus}.}
    \label{fig:vfoa}
\end{figure}

We model the field of attention as the central region of the field of view.
The field of view itself is approximated to a cone spanned from the nasal
depression (sellion) of the human face.

The dimensions of the human field of view have been measured at different level
of accuracy. Holmqvist~\cite{holmqvist2011eye} models it with an horizontal
aperture of $ \pm 40\degree $ and a vertical aperture of $ \pm 25\degree $,
while Spector~\cite{walker1980clinical} for instance suggests $60\degree$ up,
$75\degree$ down, $60\degree$ inwards (towards the nose) and $95\degree$
outwards.  Previous work on visual perspective taking for social
robotics~\cite{sisbot2011situation} simply use a cone of $30\degree$.  In this
work, we follow Holmqvist and thus model the field of view as a cone with an elliptical
cross-section of $80\degree$ by $50\degree$.

We then approximate the visual focus of attention (VFoA) of the human to the
objects which lay inside this field of view (Figure~\ref{fig:vfoa}). At a
given time, more than one object can therefore be \emph{in focus}.

Our implementation has two limitations: objects are approximated to their origin
(\ie objects are either in or out of the field of attention), and we do not
check actual visibility: one object could be hidden by another, it would still
be considered as in focus. We did not address these limitations since our
experimental setup (involving relatively small objects with no occlusions) did
not require it. Techniques for a more complete assessment of the visual
perspective can be found in~\cite{sisbot2011situation} for instance.

Within these limitations, computing if object $A(x_A,y_A,z_A)$ is in the field
of attention of the human requires first to transform the coordinates
$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
inequality (with $fov$ the aperture, and assuming that the main axis of the
field of attention is aligned on the face's $\vec{x}$):

@@

Our approach assumes that the poses of the objects of interest are available to
the system: as described in section~\ref{sec:system}, our implementation relies
on the ROS {\sc tf} framework to manage and make available to all software
modules the list of poses of existing objects (represented as {\it frames}), and
dedicated perception modules are in charged of publishing up-to-date
informations regarding the location of the points of interest (the so-called
\emph{situation assessment}). Due to the nature of the experiment, most of the
points of interest considered for the experimental validation presented
hereafter are actually taken as static with respect to the robot.

@@

As presented above, we use the (6D) head pose as an approximation of the
actual gaze direction, and we approximate from here the participant's field of
attention. The assumption that such an approximation of the field of attention
allows to derive the actual focus of attention forms the hypothesis that we
want to validate experimentally. The experiment involves child-robot
interactions in the context of handwriting remediation.
This section details the experimental procedure and presents our results.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{experimental_setup}
    \caption{\small \textbf{Experimental setup}: face-to-face interaction with a {\sc
            nao} robot. The robot writes on the tactile tablet, the child then
            corrects the robot by directly overwriting its letters on the tablet
            with a stylus. The facilitator remains next to the child to guide the work. 
            The secondary tablet allows the child to tell the robot what to
            write. The areas of interest -- corresponding to potential target of
            attention -- are circled in red with their name.}
    \label{fig:setup}
\end{figure}

@@

%An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
%camera was located at 5 cm from the center of Nao's feet and its holder was tied
%to a wood base for stability purposes. Thus, the camera's objective was 9 cm
%above and $ 40\degree $ inclination with respect to the surface of the table to
%maximize the detection during the writing time. However, Nao's camera could have
%been used, always considering the \textit{tf} frame matrix update according to
%the head orientation.

@@

Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
The study took place at school, in a separate room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other for a face-to-face interaction with a {\sc Nao} robot ($M=19.6$ min, $SD=1.58$).

The experiment is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number or a full
word), and pushes a tactile tablet (equipped with a custom application that we
developed) towards the robot; the robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pull back the tablet, corrects the robot's attempt by writing him/herself
on top or next to the robot's writing, and ``send'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until he/she
decides to train the robot on another word. In total, the children performed in
average 12.16 ($SD=2.61$) rounds of writing.

Besides, once per interaction, the robot interrupts the handwriting task to
propose to tell a story (that lasted for about 2 min), and the turn-based
hand-writing task continued afterwards. The intended purpose of the
story-telling episode was to locally trigger a different set of gaze behaviors.

Details on the rationale and implementation of this experiment can be found
in~\cite{Hood:2015}.

@@

    \path (staticposes) edge [->] (static);
    \path (static) edge [->] (tf);
    \path (input) edge [->] (attention);
    \path (attention) edge [->] (tf);
    \path (pose) edge [->]  (tf);
    \path (tf) edge [->]  (focus);
    \path (focus) edge [->]  (output);

@@

The experiment was carried with an Aldebaran's {\sc Nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent as individual frames
the possible point of interests: an object is considered to be in focus when its
frame lays within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

@@

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate, the
inter-judge agreement (Cronbach's $\alpha$) and the cosine similarity
matrix.

The cosine similarity is a more reliable measurement between the ground-truth
and the robot measure of the focus of attention as it accounts for temporal
delays.\fixme{this needs to be better justified/explained.}.

@@

The main results are reported in Table~\ref{tab:results} and
Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
Figure~\ref{fig:realExpected} further gives a concrete picture of the
ground-truth \vs computed attentional targets for subject 4 (the subject with
the least successful tracking).

\begin{table*}[h!]
    \centering
    \caption{\textbf{Attention tracking accuracy}. \emph{Head pose
    tracking} is the percentage of total time of successful detection of the
    head pose; \emph{Ground-truth/robot match} is the percentage of matching time
    between manually annotated focus of attention (ground-truth) and robot's
    computed focus of attention. Total duration: 117.6 min.}

    \begin{tabular}{p{5.5cm}cccccccc}
        \toprule
        {\bf Subject} & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        {\bf Head pose tracking} (\%) & 88.2 & 83.5 & 90.5 & 83.1 & 87.9 & 85.0 & {\bf 86.4} & {\it 3.0} \\ 
        \midrule
        {\bf Ground-truth/robot match} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
        {\bf Cronbach's $\alpha$} & 0.63 & 0.81 & 0.82 & 0.59 & 0.64 & 0.78 & {\bf0.71} & {\it0.10}\\
        {\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
        \bottomrule
    \end{tabular}
    \label{tab:results}
\end{table*}

\begin{figure}[th!]
    \centering
    \includegraphics[width=0.7\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
    using a 10 sec. window size. The correspondent numbers are shown
    in figure \ref{fig:realCaptured}.}
    \label{fig:cosinematrix}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{matches-excerpt}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}, 3.5min-long excerpt).
        In blue (top line), the focus of attention as computed by the robot;
        in orange (bottom line), the focus of attention as manually annotated
        (ground-truth). The bottom graph shows agreement between both (whenever
        the head is detected).}

@@

During the whole interaction, the head pose of the children was satisfactorily
and consistently tracked (86\% of the time in average, $SD=3.0$), which is
expected for a face-to-face interaction with a static head-tracking camera
(meaning that the child head would remain in the field of view of the camera
most of the time). This is however comforting in terms of suitability of our
approach for head pose estimation with children in field experiments.

In our experiment, the primary causes of lost head pose were occlusions with the
hands (similar to the middle-bottom picture in
Figure~\ref{head_pose_real_world}), close proximity with the tablet while
writing, and gaze directed to the facilitator (sitting directly on the left of
the child, Figure~\ref{fig:realSetup}).

In terms of attention tracking, two groups of children stand out of
Table~\ref{tab:results}: the robot's estimate of the focus of attention for
subjects 1, 4 and 5 has questionable agreement with the ground-truth (Cronbach's
$\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
($\alpha$ close to 0.8). This does not correlate with the success rate of the
head tracking itself, and, after further analysis of the videos, it rather
reflects that the first group of children rely more on their eyes to direct
their gaze than the second group. However, and notwithstanding the small sample
size, it is rather intriguing that the two groups appear so clearly separated.
This certainly calls for further experimental investigation.

@@

    \begin{algorithmic}[1]
    \Procedure{Compute With-me-ness} {}
    \State $d_w, d_e \gets 0$
    \State $t \gets t_{start}$
    \Repeat
    \If{$task(t) \neq$ nil \textbf{and} \par
    \hskip\algorithmicindent$F(task(t)) \neq \varnothing$ \textbf{and} \par
    \hskip\algorithmicindent $f(t) \neq$ nil } \label{algline:skiplosttrack}
    \If{$f(t) \in F(task(t))$}
            \State $d_w \gets d_w + 1$
        \EndIf
        \State $d_e \gets d_e + 1$
    \EndIf
    \State $t \gets t + 1$
    \Until{$t = t_{end}$}

@@

    \caption{\textbf{Computation of \emph{with-me-ness}}. $d_w$ stands for the duration
        the human is \emph{with} the robot, $d_e$ stands for the duration where
        the human would be \emph{expected to be with} the robot, $task(t)$
        represents the task performed by the robot at time $t$ (possibly none),
        $F(task)$ represents the (possibly empty) set of expected attentional
        targets associated to task $task$, $f(t)$ represents the actual focus of
        attention measured at time $t$. $\mathcal{W}_{[start, end]}$ represents
        the level of \emph{with-me-ness} from $t_{start}$ to $t_{end}$.}
    \label{alg:with-me-ness}
\end{algorithm}

>>>>

``Engagement'', the cognitive, affective and behavioral state of interaction
with a computer application that ``makes the user want to be
there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
domains. Specifically in robotics, several variables and social signals have been proposed in the
literature to quantify it. A recent review of these is presented in~\cite{ivaldi2015towards}.

@@

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
with the current state of the interaction. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
\emph{connection} events such as directed gaze, mutual facial gaze is proposed. Not
relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
a determining factor for the assessment. Dinally, a recent study with
social robots in face-to-face scenarios~\cite{anzalone} explores a set of metrics based on
non-verbal cues but they also underline possible limitations in long-term
scenarios.

The variety of these approaches reflects the fact that \emph{engagement} remains
a broad concept, fairly ill-defined and thus difficult to operationalize.
Therefore, instead of introducing ``yet another metric of engagement'', we
propose in this paper to introduce the more specific concept of
``\emph{with-me-ness}'': to what extend the human is ``with me'', the
robot, during the interaction.

We assess this level of ``with-me-ness'' by combining real-time estimation of the
attentional focus of the human with the expected, {\it a priori} focus
requested by the task, as estimated by the robot.

This article is organized in three sections: we first present a novel
method for on-line estimation of the focus of attention based on fast 6D head
pose estimation. We then propose to validate this approach with 
a real-world field experiment with children. We finally show how to compute
``with-me-ness'' over the course of the interaction to provide a new \emph{in-the-moment} measurement of
the quality of interaction, and compare levels computed by the robot with a
ground truth based on manual post-hoc annotations.

@@

The relation between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction, and the related gaze patterns, has been part of classic textbooks
like~\cite{argyle1969social} since decades. As such, there is little doubt that
measuring the direction of gaze is a useful proxy to estimate the (visual) focus
of attention of a social agent, and indeed this is one of the basic tool used in
social psychology.

Estimating attention using gaze is not new to robotics either. A
recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
way eye gaze research in HCI and social robotics. It remains however an active field of research, as illustrated by
several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
performing such a measure on a robot, in real-time, and in ecologically
valid environments (which rules out bulky or invasive apparatus like
eye-trackers, or techniques requiring fine calibration and/or static
interactions) remains a challenge in HRI.

Looking at techniques that both operate on-line and have been deployed in field
experiments, one finds that most approaches rely on head pose estimation alone
(no eye gaze tracking) and are generally based on depth sensors (RGB-D) .
Fanelli~\etal provides a overview of these approaches in~\cite{fanelli2012real},
and recent examples include~\cite{baxter2014tracking, anzalone}.

Approaches based on monocular 2D vision have been explored as
well~\cite{peters2010investigating}, with however limited robustness to
occlusions or lightning conditions, and over-reliance on tracking to maintain
real-time performances.  Our work relies on recent advances on template-based
face alignment~\cite{kazemi2014one} that allows fast (in the order of a few
milliseconds) facial features extraction on 2D images, combined with 3D model
fitting, to obtain a fast, robust and stable 6D head pose estimate, that we
successfully deployed in field experiments involving child-robot interactions.

We derive the field of attention from the head pose: this is supported by
previous work, like~\cite{stiefelhagen2002tracking} that shows that the head orientation's contribution in
overall gaze direction is 68.9\%, which further translates into a 88.7\%
accuracy in estimating the focus of attention from head pose only in a
particular meeting scenario (using eye and head tracking).

While previous preliminary research in HRI seemed on the contrary to indicate that deriving
attentional focus from head pose alone would not be accurate
enough~\cite{kennedy2015head}, we found in our case acceptable levels of
agreement between the robot observations and manual post-hoc annotations, as
detailled hereafter.

@@

As explained, we derive the visual field of attention from the head pose. Our
technique only involves a single monocular RGB camera used for facial
feature extraction, and a static simplified 3D mesh of a human head.
68 facial features are extracted using a fast
template-based face alignment algorithm by Kazemi and
Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
library~\cite{dlib09}.  Eight of these features are
then matched to their 3D counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$ algorithm
(OpenCV's implementation) to compute the translation and rotation of the head
with respect to the camera frame. With this approach, knowing the intrinsic
parameters of the camera (calibrated camera) is required for an accurate
estimation of the absolute 3D localization of the head.

Besides being fast, the face alignment algorithm has been found to perform well
in terms of robustness, including in a range of difficult situations
(occlusions, large head rotations) as encountered in field experiments.
Figure~\ref{head_pose_real_world} shows a selection of such difficult scenes with
one child.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{field_of_attention}
    \caption{\small The visual field of attention is approximated to a
        40$\degree$ cone, spanning from the head's sellion. The objects whose 3D
    pose intersect with this cone are considered \emph{in focus}.}
    \label{fig:vfoa}
\end{figure}

We model the field of attention as the central region of the field of view.  The
field of view itself is approximated to a cone spanned from the nasal depression
(sellion) of the human face. Different dimensions for the human field of view
can be found in the litterature: Holmqvist~\cite{holmqvist2011eye} models it
with an horizontal aperture of $ \pm 40\degree $ and a vertical aperture of $
\pm 25\degree $, while Spector~\cite{walker1980clinical} for instance suggests
$60\degree$ up, $75\degree$ down, $60\degree$ inwards (towards the nose) and
$95\degree$ outwards.  Previous work on visual perspective taking for social
robotics~\cite{sisbot2011situation} model the field of attention as a cone of
$30\degree$. We retained in this work a slightly wider aperture of $40\degree$.

We then approximate the visual \emph{focus} of attention (VFoA) of the human to the
objects which lay inside this field of attention (Figure~\ref{fig:vfoa}). At a
given time, more than one object can therefore be \emph{in focus}.

Our implementation has two limitations: objects are approximated to points
(their origin), and we do not
check actual visibility: one object could be hidden by another, it would still
be considered as in focus. We did not address these limitations since our
experimental setup (involving relatively small objects with no occlusions) did
not mandate it. Techniques for more accurate assessment of the visual
perspective of the human peer can be found in~\cite{sisbot2011situation} for instance.

Within these limitations, computing if object $A(x_A,y_A,z_A)$ is in the field
of attention of the human requires first to transform the coordinates
$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
simple inequality (with $fov$ the aperture, and assuming that the main axis
$\vec{x}$ of the field of attention points forward):

@@

Our approach assumes that the pose of the objects of interest are available to
the system: as described in section~\ref{sec:system}, our implementation relies
on the ROS {\sc tf} framework to manage and make available to all software
modules the list of poses of existing objects (represented as {\it frames}), and
dedicated perception modules are in charged of publishing up-to-date
informations regarding the location of the objects of interest (the so-called
\emph{situation assessment}). Due to the nature of the experiment, most of the
points of interest considered for the experimental validation presented
hereafter are static with respect to the robot, thus simplifying the scene
perception.

@@

As presented above, we use the 6D head pose as an approximation of the
actual gaze direction, and we further approximate from here the participant's field of
attention. The assumption that such an double approximation of the field of attention
allows to derive the actual focus of attention needs to validated experimentally.
Our proposed experiment involves child-robot
interactions in the context of handwriting remediation.
This section details the experimental procedure and presents our results.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{experimental_setup}
    \caption{\small \textbf{Experimental setup}: face-to-face interaction with a {\sc
            nao} robot. The robot writes on the tactile tablet, the child then
            corrects the robot by directly overwriting the words on the tablet
            with a stylus. The facilitator remains next to the child to guide the work. 
            The secondary tablet allows the child to tell the robot what to
            write. The areas of interest -- corresponding to potential target of
            attention -- are circled in red with their name.}
    \label{fig:setup}
\end{figure}

@@

%An RGB camera has been used to acquire images of 640x480 pixels at 25 fps. The
%camera was located at 5 cm from the center of nao's feet and its holder was tied
%to a wood base for stability purposes. Thus, the camera's objective was 9 cm
%above and $ 40\degree $ inclination with respect to the surface of the table to
%maximize the detection during the writing time. However, nao's camera could have
%been used, always considering the \textit{tf} frame matrix update according to
%the head orientation.

@@

Six children (ages 5 to 6, 3 boys, 3 girls) were enrolled for this study.
The study took place at school, in a separate room (the computer
lab). The participants were chosen by the teacher, and would come one after the
other to interact with the robot (duration: $M=19.6$ min, $SD=1.58$).

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number or a full
word), and presents a tactile tablet (equipped with a custom writing
application) to the robot; the robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pull back the tablet, corrects the robot's attempt by writing him/herself
on top or next to the robot's writing, and ``send'' his/her demonstration to the
robot by pressing a small button on the tablet. The robot learns from this
demonstration and tries again. The child continues the turn-taking until he/she
decides to train the robot on another word. In total, the children performed in
average 12.16 ($SD=2.61$) rounds of writing (complete details on the rationale
and implementation of this experiment can be found in~\cite{Hood:2015}).

Besides, once per interaction, the robot interrupts the handwriting task to
propose to tell a story (taking about 2 min), and the turn-based
hand-writing task continues afterwards. The intended purpose of the
story-telling episode is to locally trigger a different set of attention behaviors.

@@

    \path (staticposes) edge [->] (static);
    \path (static) edge [->] node[label,below] {objects poses} (tf);
    \path (input) edge [->] (attention);
    \path (attention) edge [->] node[label,below] {child face} (tf);
    \path (pose) edge [->] node[label,right] {robot pose} (tf);
    \path (tf) edge [->]  (focus);
    \path (focus) edge [->]  (output);

@@

The experiment was carried with an Aldebaran's {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent as individual frames
the possible point of interests: an object is considered to be in focus when its
frame lays within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

@@

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate and the
inter-judge agreement (Cronbach's $\alpha$).
%and the cosine similarity
%matrix.

%The Cronbach's $\alpha$ can be defined as an average of the correlations between 
%the items or values at the same time step \textit{t}. Therefore, results can vary 
%significantly if both sets of values are time shifted. However,
%the cosine similarity becomes a useful measurement in this case as it accounts 
%for temporal delays. Specifically, the ground-truth values acquired though human 
%annotations may not match the measurements of the focuses of attention computed 
%by the robot at the exactly same time step \textit{t} even though the correspondent
%curves may look very similar in shape.

@@

\begin{table*}[ht!]
    \centering
    \caption{\textbf{Attention tracking accuracy}. \emph{Head pose
    tracking} is the percentage of total time of successful detection of the
    head pose; \emph{Agreement} is the percentage of matching time
    between manually annotated focus of attention (ground-truth) and robot's
    computed focus of attention. Total duration: 117.6 min.}

    \begin{tabular}{p{5.5cm}cccccccc}
        \toprule
        {\bf Subject} & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        {\bf Head pose tracking} (\%) & 88.2 & 83.5 & 90.5 & 83.1 & 87.9 & 85.0 & {\bf 86.4} & {\it 3.0} \\ 
        \midrule
        {\bf Agreement} (\%) & 61.1 & 85.5 & 80.8 & 62.2 & 63.4 & 83.9 & {\bf 72.8} & {\it 11.7}\\
        {\bf Cronbach's $\alpha$} & 0.63 & 0.81 & 0.82 & 0.59 & 0.64 & 0.78 & {\bf0.71} & {\it0.10}\\
        %{\bf Cosine affinity} & 0.73 & 0.82 & 0.87 & 0.73 & 0.77 & 0.89 & {\bf 0.8} & {\it 0.07} \\
        \bottomrule
    \end{tabular}
    \label{tab:results}
\end{table*}

The main results are reported in Table~\ref{tab:results}
%and Figure~\ref{fig:cosinematrix} (cosine similarity matrix);
Figure~\ref{fig:realExpected} further gives a concrete picture of the
ground-truth \vs computed attentional targets for subject 4 (the subject with
the \emph{least} successful tracking).

%
%\begin{figure}[th!]
%    \centering
%    \includegraphics[width=0.7\columnwidth]{bitmap}
%    \caption{\small Cosine similarity matrix between ground-truth and robot-measured VFoA
%    using a 10 sec. window size. The correspondent numbers are shown
%    in figure \ref{fig:realCaptured}.}
%    \label{fig:cosinematrix}
%\end{figure}

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=\linewidth]{matches-excerpt}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}, 3.5min-long excerpt).
        In blue (top line), the focus of attention as computed by the robot;
        in orange (bottom line), the focus of attention as manually annotated
        (ground-truth). The bottom graph shows agreement between both (whenever
        the head is detected).}

@@

During the whole interaction, the head pose of the children was satisfactorily
and consistently tracked (86\% of the time in average, $SD=3.0$). While this
high score is expected for a face-to-face interaction with a static head-tracking camera
(meaning that the child head would remain in the field of view of the camera
most of the time), this is still comforting in terms of suitability of our
approach for head pose estimation with children in field experiments of this
kind. Expectedly, the primary causes of lost head pose were occlusions with the
hands (similar to the middle-bottom picture in
Figure~\ref{head_pose_real_world}), close proximity with the tablet while
writing, and gaze directed to the facilitator (who was sitting directly on the left of
the child, Figure~\ref{fig:realSetup}).

In terms of attention tracking, two groups of children stand out of
Table~\ref{tab:results}: the robot's estimate of the focus of attention for
subjects 1, 4 and 5 has questionable agreement with the ground-truth (Cronbach's
$\alpha$ close to 0.6), while for subjects 2, 3 and 6, we find a good agreement
($\alpha$ close to 0.8). After further analysis of the videos, it
reflects that the first group of children rely more on their eyes to direct
their gaze than the second group. Note that, notwithstanding the small sample
size, it is rather intriguing that the two groups appear so well distinct.
This calls for further experimental investigation.

@@

    \begin{algorithmic}[1]
    \Procedure{Compute With-me-ness} {}
    \State $d_w, d_e \gets 0$
    \State $t \gets t_{start}$
    \Repeat
    \If{$task(t) \neq$ nil \textbf{and} \par
    \hskip\algorithmicindent$F(task(t)) \neq \varnothing$ \textbf{and} \par
    \hskip\algorithmicindent $f(t) \neq$ nil } \label{algline:skiplosttrack}
    \If{$f(t) \in F(task(t))$}
            \State $d_w \gets d_w + \delta_t$
        \EndIf
        \State $d_e \gets d_e + \delta_t$
    \EndIf
    \State $t \gets t + \delta_t$
    \Until{$t = t_{end}$}

@@

    \caption{\textbf{Computation of \emph{with-me-ness}}. $d_w$ stands for the duration
        the human is actually \emph{with} the robot, while $d_e$ stands for the
        total time where the human would be \emph{expected to be with} the robot, $task(t)$
        represents the task performed by the robot at time $t$ (possibly none),
        $F(task)$ represents the (possibly empty) set of expected attentional
        targets associated to task $task$, $f(t)$ represents the actual focus of
        attention of the human measured at time $t$. $\mathcal{W}_{[start, end]}$ represents
        the level of \emph{with-me-ness} from $t_{start}$ to $t_{end}$.}
    \label{alg:with-me-ness}
\end{algorithm}

Algorithm~\ref{alg:with-me-ness} provides a formal way of computing the level of
with-me-ness $\mathcal{W}$ between two time points $[t_{start}, t_{end}]$. A
notable difference with the original definition by Sharma~\etal is that, at a
given time $t$, the task $task(t)$ performed by the robot may elicit more than
one attentional target; thus, at a given time, more than one location can be
regarded as possible \emph{expected} focuses of attention for the human. For
example, a robot which is writing, could typically elicit gazes to its hand as
well as to its head. A human looking at either of these locations would be
considered to be \emph{with} the robot in terms of
interaction\footnote{Considering a probabilistic model of
attention expectations would be an interesting extension of this metric.}. Also notable, we exclude from the computation of
$\mathcal{W}$ all of the periods of time where the user's focus of attention can
not be estimated (typically because the user's
face is not visible at those times).
